### 2015-03-17 Kokkos_PG_MachineModel.tex
Machine model: Filled out the section on "Program execution."

@@

@@

Kokkos imposes a very weak consistency model.
For example, it does \emph{not} assume sequential consistency.
Much like MPI 3 one-sided communication
or other partitioned global address space (PGAS) programming models,
Kokkos requires an execution space to issue a \emph{memory fence}
to a given memory space, 
in order to bring that memory space into a consistent state.
The fence forces completion of all outstanding memory writes,
and reorders all loads to occur after all those writes complete.

\section{Asynchronous parallel dispatch}\label{S:Model:Async}

Kokkos lets users tell execution spaces to execute parallel operations.
These include parallel for, reduce, and scan (see Chapter \ref{C:Dispatch})
as well as View initialization (see Chapter \ref{C:View}).
We name the class of all such operations \emph{parallel dispatch}.

In Kokkos, parallel dispatch executes \emph{asynchronously}.  
This means that they may return ``early,'' before they have actually completed.
However, despite this asynchrony,
they execute in sequence with respect to other Kokkos operations.
This matters for things like timing.
For example, a \lstinline!parallel_for! may return ``right away,''
so if you want to measure how long it takes,
you must first call \lstinline!fence()! on that execution space.
This forces all functors to complete before \lstinline!fence()! returns.

\section{Atomic updates}\label{S:Model:Atomic}

>>>>

\subsection{Types and instances of spaces}\label{SS:Model:Spaces:Types}

\subsection{Analogy with operating system processes}\label{SS:Model:Spaces:Analogies}

@@

\section{Thread teams}\label{S:Model:Teams}

Kokkos supports multiple levels of parallelism.
In particular, it groups threads into \emph{teams}.
A \emph{thread team} is a collection of one or more parallel ``threads'' of execution.
Kokkos allows an arbitrary number of teams -- the \emph{league size}.
Hardware constrains the number of threads in a team -- the \emph{team size}.

Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
and share a ``scratch pad'' memory which they may use for temporary storage.
Scratch pad memory exists only during parallel operations;
allocations in it do not persist across kernels.
Teams themselves may run in any order,
and may not necessarily run all in parallel.
For example, if the user asks for $T$ teams,
the hardware may choose to run them one after another in sequence,
or in groups of up to $G$ teams at a time in parallel.

Users may \emph{nest} parallel operations.
Teams may perform one parallel operation (for, reduce, or scan),
and threads within each team may perform another, possibly different parallel operation.
Different teams may do entirely different things.
For example, all the threads in one team may execute a \lstinline!parallel_for!,
and all the threads in a different team may execute a \lstinline!parallel_scan!.
Different threads within a team may also do different things.
However, performance may vary if threads in a team ``diverge'' in their behavior
(e.g., take different sides of a branch).
Chapter \ref{C:Hierarchical} shows how the C++ implementation of Kokkos exposes thread teams.

NVIDIA's CUDA programming model inspired Kokkos' thread team model.
The scratch pad memory corresponds with CUDA's per-team ``shared memory.''
The ``league / team'' vocabulary comes from OpenMP 4.0,
and has many aspects in common with our thread team model.
We have found that programming to this model results in good performance,
even on computer architectures which only implement parts of the full model.
For example, most multicore processors in common use for high-performance computing lack ``scratch pad'' hardware.
However, if users request a scratch pad size that fits comfortably in the largest cache shared by the threads in a team,
programming as if a scratch pad exists forces users to address locality in their algorithms.
This also reflects the common experience that rewriting a code for more restrictive hardware,
then porting the code \emph{back} to conventional hardware,
tends to improve performance relative to an unoptimized code.

\section{Program execution}\label{S:Model:Exec}

It is tempting to try to define formally what it means for a processor to execute code.
None of us authors have a background in logic or what computer scientists call ``formal methods,''
so our attempt might not go very far!
We will stick with informal definitions and rely on Kokkos' C++ implementation
as an existence proof that the definitions make sense.

Kokkos lets users tell execution spaces to execute parallel operations.
These include parallel for, reduce, and scan (see Chapter \ref{C:Dispatch})
as well as View allocation and initialization (see Chapter \ref{C:View}).
We name the class of all such operations \emph{parallel dispatch}.

From our perspective, there are three kinds of code:
\begin{enumerate}
\item Code executing inside of a Kokkos parallel operation
\item Code outside of a Kokkos parallel operation that asks Kokkos to
  do something (e.g., parallel dispatch itself)
\item Code that has nothing to do with Kokkos
\end{enumerate}
The first category is the most restrictive.
Section \ref{S:Model:Teams} above explains restrictions on inter-team synchronization.
In general, we limit the ability of Kokkos-parallel code to invoke Kokkos operations
(other than for nested parallelism; 
see Section \ref{S:Model:Teams} above and Chapter \ref{C:Hierarchical}).
We also forbid dynamic memory allocation (other than from the team's scratch pad) in parallel operations.
Whether Kokkos-parallel code may invoke operating system routines or third-party libraries
depends on the execution and memory spaces being used.
Regardless, restrictions on inter-team synchronization have implications for things like filesystem access.

\emph{Kokkos threads are for computing in parallel,}
not for overlapping I/O and computation,
and not for making graphical user interfaces responsive.
Use other kinds of threads (e.g., operating system threads) for the latter two purposes.
You may be able to mix Kokkos' parallelism with other kinds of threads;
see Section \ref{SS:Model:Exec:ThreadSafety}.
Kokkos' developers are also working on a task parallelism model
that will work with Kokkos' existing data-parallel constructs.

\subsection{Reproducible reductions and scans}\label{SS:Model:Exec:Repro}

Kokkos promises \emph{nothing} about the order in which the iterations of a parallel loop occur.
However, it \emph{does} promise that if you execute the same parallel reduction or scan,
using the same hardware resources and run-time settings,
then you will get the same results each time you run the operation.
``Same results'' even means ``with respect to floating-point rounding error.''

\subsection{Asynchronous parallel dispatch}\label{SS:Model:Exec:Async}

This concerns the second category of code, that calls Kokkos operations.
In Kokkos, parallel dispatch executes \emph{asynchronously}.  
This means that it may return ``early,'' before it has actually completed.
Nevertheless, it executes \emph{in sequence} with respect to other Kokkos operations on the same execution or memory space.
This matters for things like timing.
For example, a \lstinline!parallel_for! may return ``right away,''
so if you want to measure how long it takes,
you must first call \lstinline!fence()! on that execution space.
This forces all functors to complete before \lstinline!fence()! returns.

\subsection{Thread safety?}\label{SS:Model:Exec:ThreadSafety}

Users may wonder about ``thread safety,'' that is,
whether multiple operating system threads may safely call into Kokkos concurrently.
Kokkos' thread safety depends on both its implementation, 
and on the execution and memory spaces that the implementation uses.
The C++ implementation has made great progress towards (non-Kokkos) thread safety of View memory management.
For now, however, the most portable approach is for only one (non-Kokkos) thread of execution to control Kokkos.
Also, be aware that operating system threads might interfere with Kokkos' performance,
depending on the execution space that you use.

Thread safety with respect to Kokkos' threads is a different matter.
View (see Chapter \ref{C:View}) promises safety of its memory management.
That is, you may safely access Views inside of Kokkos' parallel operations.
Kokkos provides some synchronization constructs as well; see the following section.

@@

Kokkos imposes a very weak consistency model across an entire execution space.
For example, it does \emph{not} assume sequential consistency.
Much like MPI 3 one-sided communication
or other partitioned global address space (PGAS) programming models,
Kokkos requires an execution space to issue a \emph{memory fence}
to a given memory space, 
in order to bring that memory space into a consistent state.
The fence forces completion of all outstanding memory writes,
and reorders all loads to occur after all those writes complete.

\subsection{Atomic updates}\label{SS:Model:Consistency:Atomic}
